\Large\textbf{General Stage Untangling with Applications to Algorithm Derivation}

### 2014-03-07 paper/paper.tex
Fixed up first stage splitting for let and if, in the theory.
Added rules for sums and functions.

\title{\Large\textbf{General Stage Untangling with Applications to Algorithm Derivation}}
\author{Nicolas Feltman et al.}
\begin{document}
\maketitle
\section{Introduction}

Many computational tasks are premised on the notion of multiple domains, places and times at which computation occurs.  Yet programming languages traditionally view their execution environments as monolithic, specifying the What and the How, but not the When nor the Where.   As always, where the language falls short, a programmer instead must bear the burden.  Inevitably, this comes at the cost of expressivity, safety, or both.  In this work, we present a language for specifying certain multi-domain computations, sacrificing neither the expresivity nor the safety that programmers have come to expect in single-domain languages.

Consider the programming of a pipeline system.  Generally in such a system, data comes in to one processor, is transformed to an intermediate result, and then fed to later processors, possibly with some buffering, duplication, and reording in between.  We can view an algorithm for such as system as one computation, split accross two spatial domains (processors).  For a concrete example, we draw inspiration from graphics systems.  Lets say that we want to map a function $f : \mathrm{PixelCoord} \to \mathrm{Color}$ over every pixel on a 1024x768 screen.   Due to geometric coherence inherent in the problem, it's often advantageous to compute some information at the level of coarse 32x32 tiles (e.g. a list of the scene primitives which interact with that part of the screen) whose results can be used to accelerate the final per-pixel calculation.  In adition to yielding algorithmic benefits, this decomposition strategy maps well to parallel hardware.

@@

We'll start by analyzing a simple two-stage language called \corelang.  We say it is simple in that it supports boolean, integer, and product types, as well as let bindings, if expressions, and various primitive operations.  Notably, we leave out functions, fixed point operators, and full sum types, which will be added later.

The grammar for types, terms, values, and contexts in \corelang are given in Figure \ref{fig:coreSyntax}.  The first thing to notice is that stage $\bbone$ and stage $\bbtwo$ code are syntactically seperated at the type, term, and value levels, despite heavy overlap.  It would be tempting to merge the two stages into one syntactic class and maintain the stage invariants with the typing judgements.  Such a trick would reduce the total number of rules presented later, but at the cost of making those rules type derivation-directed, rather than syntax directed.  Alternatively, we chose to have one context merged accross both stages, rather than one context for each stage.  The intended effect of this decision is for bindings in one stage to shadow binding to the same variables in the other stage.  The reasons for this will become clear later.

@@

\begin{figure}
\caption{\corelang~Syntax}
\label{fig:coreSyntax}
\centering
\begin{tabular}{ll} 
$\begin{aligned}
\typeo &::= \text{unit}~|~\text{int}~|~\text{bool} \\
&\gbar \typeo \times \typeo \\
&\gbar \fut \typet 
\end{aligned} $  
& 
$\begin{aligned}
\typet &::=  \text{unit}~|~\text{int}~|~\text{bool} \\
&\gbar \typet \times \typet \\
&
\end{aligned} $  
\\ 
$\begin{aligned}
\expro &::= ()~|~\inte~|~\bool  \\
&\gbar \letin{\var}{\expro}{\expro} \\
&\gbar \var \\
&\gbar (\expro, \expro) \\
&\gbar \pi_1~\expro \gbar \pi_2~\expro \\
&\gbar \ifthen {\expro}{\expro}{\expro} \\
&\gbar \next~\exprt 
\end{aligned} $ 
& 
$\begin{aligned}
\exprt &::= ()~|~\inte~|~\bool \\
&\gbar \letin{\var}{\exprt}{\exprt} \\
&\gbar \var \\
&\gbar (\exprt, \exprt) \\
&\gbar \pi_1~\exprt \gbar \pi_2~\exprt \\
&\gbar \ifthen {\exprt}{\exprt}{\exprt} \\
&\gbar \prev~\expro
\end{aligned} $
\\ 
$\begin{aligned}
\valo &::= ()~|~\inte~|~\bool \\
&\gbar \valprod {\valo} {\valo} \\
&\gbar \valnext~\valt
\end{aligned} $
& 
$\begin{aligned}
\valt &::= ()~|~\inte~|~\bool \\
&\gbar \valprod {\valt} {\valt} \\
& 
\end{aligned} $
\\
$\begin{aligned}
\contextot &::=\emptyC \\
&\gbar \contextot, \var : \typeo ^\bbone \\
&\gbar \contextot, \var : \typet ^\bbtwo
\end{aligned} $
\end{tabular}
\end{figure}

@@

The static semantics for \corelang~is given in Figure \ref{fig:coreStatics}.  Modulo the two-stage restriction and current absence of functions, it's based off of the system in \cite{davies96}, which was inspired by temporal logic (the source of our $\fut$ modality).  The static semantics comprises two judgements.  Those are:

@@

\begin{figure}
\caption{\corelang~Static Semantics}
\label{fig:coreStatics}
\begin{mathpar}
\infertypesone [\rmunit] 	{\ty {()}\rmunit}						{\cdot} \and
\infertypesone [int]			{\ty {i}{\rm int}}					{\cdot} \and
\infertypesone [bool]			{\ty {b} \rmbool}						{\cdot} \and
\infertypesone [hyp]			{\ty x A}								{x : A^\bbone \in \Gamma} \and
\infertypestwo [\rmunit]		{\ty {()}{\rmunit}}					{\cdot} \and
\infertypestwo [int]			{\ty {i}{\rm int}}					{\cdot} \and
\infertypestwo [bool]			{\ty {b} \rmbool}						{\cdot} \and
\infertypestwo [hyp]			{\ty x A}								{x : A^\bbtwo \in \Gamma} \and
\infertypesone [let]			{\ty {\letin{x}{e_1}{e_2}}{B}}		{\ty {e_1} A & \ty [\Gamma,x:A^\bbone] {e_2} B} \and
\infertypestwo [let]			{\ty {\letin{x}{e_1}{e_2}}{B}}		{\ty {e_1} A & \ty [\Gamma,x:A^\bbtwo] {e_2} B} \and
\infertypesone [\times I]	{\ty {(e_1,e_2)}{A\times B}}			{\ty {e_1} A & \ty {e_2} B} \and
\infertypesone [\times E_1]	{\ty {\pi_1~e} A}						{\ty e {A\times B}} \and
\infertypesone [\times E_2]	{\ty {\pi_2~e} B}						{\ty e {A\times B}} \and
\infertypestwo [\times I]	{\ty {(e_1,e_2)}{A\times B}}			{\ty {e_1} A & \typestwo {e_2} B} \and
\infertypestwo [\times E_1]	{\ty {\pi_1~e} A}						{\ty e {A\times B}} \and
\infertypestwo [\times E_2]	{\ty {\pi_2~e} B}						{\ty e {A\times B}} \and
\infertypesone [if] 			{\ty {\ifthen{e_1}{e_2}{e_3}} A}	{\ty {e_1} \rmbool & \ty {e_2} A & \ty {e_3} A} \and
\infertypestwo [if]			{\ty {\ifthen{e_1}{e_2}{e_3}} A}	{\ty {e_1} \rmbool & \ty {e_2} A & \ty {e_3} A} \and
\infertypesone [\fut I]		{\ty {\next~e}{\fut A}}				{\typestwo e A} \and
\infertypestwo [\fut E]		{\ty {\prev~e} A}						{\typesone e {\fut A}} \and
\end{mathpar}
\end{figure}

@@

The dynamics semantics for \corelang~is given in Figure \ref{fig:erasureSemantics} in big-step style.  Predictably, it comprises two judgements:

@@

\begin{figure}
\caption{\corelang~Erasure Dynamic Semantics}
\label{fig:erasureSemantics}
\begin{mathpar}
\inferreduceone [\rmunit]		{\red {()}{()}}							{\cdot} 											\and
\inferreduceone [int]				{\red {i}{i}}								{\cdot} 											\and
\inferreduceone [bool]			{\red {b}{b}}								{\cdot} 											\and
\inferreduceone [let]				{\red {\letin{x}{e_1}{e_2}}{v_2}}		{\sub [1] & \red{[v_1/x]e_2}{v_2}} 				\and
\inferreducetwo [\rmunit]		{\red {()}{()}}							{\cdot} 											\and
\inferreducetwo [int]				{\red {i}{i}}								{\cdot} 											\and
\inferreducetwo [bool]			{\red {b}{b}}								{\cdot} 											\and
\inferreducetwo [let]				{\red {\letin{x}{e_1}{e_2}}{v_2}}		{\sub [1] & \red{[v_1/x]e_2}{v_2}} 				\and
\inferreduceone [\times I]		{\red {(e_1,e_2)}{\valprod{v_1}{v_2}}}	{\sub [1] & \sub [2]} 							\and
\inferreduceone [\times E_1]		{\red {\pi_1~e}{v_1}}						{\red{e}{\valprod{v_1}{v_2}}} 					\and
\inferreduceone [\times E_2]		{\red {\pi_2~e}{v_2}}						{\red{e}{\valprod{v_1}{v_2}}} 					\and
\inferreducetwo [\times I]		{\red {(e_1,e_2)}{\valprod{v_1}{v_2}}}	{\sub [1] & \sub [2]}								\and
\inferreducetwo [\times E_1]		{\red {\pi_1~e}{v_1}}						{\red{e}{\valprod{v_1}{v_2}}} 					\and
\inferreducetwo [\times E_2]		{\red {\pi_2~e}{v_2}}						{\red{e}{\valprod{v_1}{v_2}}} 					\and
\inferreduceone [if_T] 			{\red {\ifthen{e_1}{e_2}{e_3}}{v}}		{\red{e_1}{\bf true}  & \red{e_2}{v}} 			\and
\inferreduceone [if_F] 			{\red {\ifthen{e_1}{e_2}{e_3}}{v}}		{\red{e_1}{\bf false} & \red{e_3}{v}}			\\
\inferreducetwo [if_T] 			{\red {\ifthen{e_1}{e_2}{e_3}}{v}}		{\red{e_1}{\bf true}  & \red{e_2}{v}} 			\and
\inferreducetwo [if_F] 			{\red {\ifthen{e_1}{e_2}{e_3}}{v}}		{\red{e_1}{\bf false} & \red{e_3}{v}}			\\
\inferreduceone [\fut I]			{\red {\next~e}{\valnext~v}}				{\reducetwosub}									\and
\inferreducetwo [\fut E]			{\red {\prev~e} v}						{\reduceone e {\valnext~v}} 
\end{mathpar}
\end{figure}

\begin{figure}
\caption{\corelang~Diagonal Semantics}
\label{fig:refSemantics}
\begin{mathpar}
\inferRedsplitone [\rmunit]		{\red {()}{()}{()}}												{\cdot} 												\and
\inferRedsplitone [int]			{\red {i}{i}{()}}													{\cdot} 												\and
\inferRedsplitone [bool]			{\red {b}{b}{()}}													{\cdot} 												\and
\inferRedsplitone [hyp]			{\red {[u,x]}{u}{x}}												{\cdot}													\and
\inferSpeculation [\rmunit]		{\red {()}{()}}													{\cdot} 												\and
\inferSpeculation [int]			{\red {i}{i}}														{\cdot} 												\and
\inferSpeculation [bool]			{\red {b}{b}}														{\cdot} 												\and
\inferSpeculation [hyp]			{\red {x}{x}}														{\cdot}													\and
\inferRedsplitone [let]			{\red {\letin{x}{e_1}{e_2}}{u_2}{\letin {x}{r_1}{r_2}}}		{\sub [1] & \red{[[u_1,x]/x]e_2}{u_2}{r_2}} 		\and
\inferSpeculation [let]			{\red {\letin{x}{e_1}{e_2}}{\letin{x}{r_1}{r_2}}}				{\sub [1] & \sub [2]}									\and
\inferRedsplitone [\times I]		{\red {(e_1,e_2)}{\valprod{u_1}{u_2}}{\valprod{r_1}{r_2}}}	{\sub [1] & \sub [2]}									\and
\inferRedsplitone [\times E_1]	{\red {\pi_1~e}{u_1}{\pi_1 r}}									{\red {e}{\valprod{u_1}{u_2}}{r}}					\and
\inferRedsplitone [\times E_2]	{\red {\pi_2~e}{u_2}{\pi_2 r}}									{\red {e}{\valprod{u_1}{u_2}}{r}} 					\and
\inferSpeculation [\times I]		{\red {(e_1,e_2)}{(r_1,r_2)}}									{\sub [1] & \sub [2]}									\and
\inferSpeculation [\times E_1]	{\red {\pi_1~e}{\pi_1 r}}										{\red {e}{r}}											\and
\inferSpeculation [\times E_2]	{\red {\pi_2~e}{\pi_2 r}}										{\red {e}{r}} 											\and
\inferRedsplitone [if_T] 		{\red {\ifthen{e_1}{e_2}{e_3}}{u}{(r_1;r_2)}}					{\red{e_1}{\bf true }{r_1} & \red{e_2}{u}{r_2}}	\and
\inferRedsplitone [if_F] 		{\red {\ifthen{e_1}{e_2}{e_3}}{u}{(r_1;r_3)}}					{\red{e_1}{\bf false}{r_1} & \red{e_3}{u}{r_3}}	\and
\inferSpeculation [if] 			{\red {\ifthen{e_1}{e_2}{e_3}}{\ifthen{r_1}{r_2}{r_3}}}		{\sub [1] & \sub [2] & \sub [3]}					\\
\inferRedsplitone [\fut I]		{\red {\next~e}{()}{r}}											{\rstwosub}											\and
\inferSpeculation [\fut E]		{\red {\prev~e}{r}}												{\rsone {e}{()}{r}}
\end{mathpar}
\end{figure}

@@

We now cover splitting, the process by which mixed code is teased apart into two individual chunks of code, one for each stage.  To begin, we first introduce the target lagnuage (for both stage), \coremono.   The grammar for \coremono~is given in figure \ref{fig:baseSyntax}.  We see that \coremono~is (as one would hope) a traditional single-stage language, with all the same basetypes as \corelang, in addition to products, if expressions, and full sums (injections and case expressions).  The presence of this last feature is a bit curious, since \corelang~did not support full sums.  The necessity for sums should become apparent in time.  The static and dynamic semantics of \coremono~are not given, but should be obvious.  They are given by the judgements:

@@

\begin{figure}
\caption{\coremono~Syntax}
\label{fig:baseSyntax}
\centering
\begin{tabular}{ll} 
$\begin{aligned}
\expr &::= ()~|~\inte~|~\bool \\
&\gbar \letin{\var}{\expr}{\expr} \\
&\gbar \var \\
&\gbar (\expr, \expr) \\
&\gbar \pi_1~\expr \gbar \pi_2~\expr \\
&\gbar \inl~\expr \gbar \inr~\expr \\
&\gbar \ifthen {\expr}{\expr}{\expr}  \\
&\gbar \caseof {\expr}{x_1.\expr}{x_2.\expr} 
\end{aligned} $
& 
$\begin{aligned}
\type &::=  \rmunit~|~\text{int}~|~\text{bool} \\
&\gbar \type \times \type  \\
&\gbar \type + \type 
\\
\val &::= ()~|~\inte~|~\bool \\
&\gbar \valprod {\val} {\val}  \\
&\gbar \valinl~\expr \gbar \valinr~\expr
\\
\context &::= \emptyC \\
&\gbar \context, \var : \type
\end{aligned} $
\end{tabular}
\end{figure}

@@

\begin{figure*}
\caption{Term Splitting}
\label{fig:termSplit}
\newcommand {\projbind} [1] {\letin{l_{#1}}{\pi_{#1} l}{r_{#1}}}
\begin{mathpar}
\infersplitone [unit]			{\spl {()}{\rmunit}{((),()),\_.()}}															{\cdot} \and
\infersplittwo [unit]			{\spl {()}{\rmunit}{(),\_.()}}																{\cdot} \and
\infersplitone [int]			{\spl {i}{\rm int}{(i,()),\_.()}}															{\cdot} \and
\infersplittwo [int]			{\spl {i}{\rm int}{(),\_.i}}																	{\cdot} \and
\infersplitone [\times I] 	{\spl {(e_1,e_2)}{A\times B}
									{ \begin{array}{l}
									\left(
										\letin{y_1}{c_1}{\letin{y_2}{c_2}{
										((\pi_1 y_1, \pi_1 y_2),(\pi_2 y_1, \pi_2 y_2))}}
									\right), \\
									l.(\projbind 1,\projbind 2)
									\end{array}}}																				{\sub 1 A & \sub 2 B} \and
\infersplittwo [\times I]	{\spl {(e_1,e_2)}{A\times B}{(p_1,p_2),l.(\projbind 1,\projbind 2)}}					{\sub 1 A & \sub 2 B} \and
\infersplitone [\times E_1]	{\spl {\pi_1~e}{A}{\letin{y}{c}{(\pi_1(\pi_1 y),\pi_2 y)},l.\pi_1~r}}					{\sub {} {A\times B}} \and
\infersplitone [\times E_2]	{\spl {\pi_2~e}{B}{\letin{y}{c}{(\pi_2(\pi_1 y),\pi_2 y)},l.\pi_2~r}}					{\sub {} {A\times B}} \and
\infersplittwo [\times E_1]	{\spl {\pi_1~e}{A}{p,l.\pi_1~r}}															{\sub {} {A\times B}} \and
\infersplittwo [\times E_2]	{\spl {\pi_2~e}{B}{p,l.\pi_2~r}}															{\sub {} {A\times B}} \and
\infersplitone [hyp]			{\spl {v}{A}{(x,()),\_.()}}																	{x:A^\bbone \in \Gamma} \and
\infersplittwo [hyp]			{\spl {v}{A}{(),\_.x}}																		{x:A^\bbtwo \in \Gamma} \\
\infersplitone [let]			{\spl {\letin{x}{e_1}{e_2}}{B}{\cdots}}														{\sub 1 A & \sub [\Gamma,x:A^\bbone] 2 B} \and
\infersplittwo [let]			{\spl {\letin{x}{e_1}{e_2}}{B}{(p_1,p_2), l.\letin{x}{(\projbind 1)}{\projbind 2}}}	{\sub 1 A & \sub [\Gamma,x:A^\bbtwo] 2 B} \and
\infersplitone [\fut I]		{\spl {\next~e}{\fut A}{((),p),l.r}}														{\splittwosub {} A} \and
\infersplittwo [\fut E]		{\spl {\prev~e}{A} {\pi_2~c,l.r}}															{\splitonesub {} {\fut A}} \and
\infersplitone [if]			{\spl {\left( \tallif {e_1}{e_2}{e_3} \right)}{A} 
								{\begin{array}{l}
									\left(\tallif {\pi_1 c_1}
										{\letin{y}{c_2}{(\pi_1 y, \inl(\pi_2 y))}}
										{\letin{y}{c_3}{(\pi_1 y, \inr(\pi_2 y))}}
									\right), \\
									l.(\caseof{l}{l_2.r_2}{l_3.r_3})
								\end{array}}}																					{\sub 1 \rmbool & \sub 2 A & \sub 3 A } \and
\infersplittwo [if]			{\spl {\left(\tallif {e_1}{e_2}{e_3}\right)}{A} 
								{(p_1,p_2,p_3),l.\left(\tallif{(\projbind 1)}{(\projbind 2)}{(\projbind 3)}\right)}}	{\sub 1 \rmbool & \sub 2 A & \sub 3 A } \and
\end{mathpar}
\end{figure*}

@@

\section {Implementation}

Not much to discuss.  It exists.  Talk about how crufty the naive transformation is.

\section{Related Work in Stage Seperation}

>>>>

\title{\Large\textbf{General Stage Untangling with Applications to Algorithm Derivation}}
\author{Nicolas Feltman et al.}
\begin{document}
\maketitle

\input {grammars}

\section{Introduction}

Many computational tasks are premised on the notion of multiple domains, places and times at which computation occurs.  Yet programming languages traditionally view their execution environments as monolithic, specifying the What and the How, but not the When nor the Where.   As always, where the language falls short, a programmer instead must bear the burden.  Inevitably, this comes at the cost of expressiveness, safety, or both.  In this work, we present a language for specifying certain multi-domain computations, sacrificing neither the expressiveness nor the safety that programmers have come to expect in single-domain languages.

Consider the programming of a pipeline system.  Generally in such a system, data comes in to one processor, is transformed to an intermediate result, and then fed to later processors, possibly with some buffering, duplication, and reordering in between.  We can view an algorithm for such as system as one computation, split across two spatial domains (processors).  For a concrete example, we draw inspiration from graphics systems.  Lets say that we want to map a function $f : \mathrm{PixelCoord} \to \mathrm{Color}$ over every pixel on a 1024x768 screen.   Due to geometric coherence inherent in the problem, it's often advantageous to compute some information at the level of coarse 32x32 tiles (e.g. a list of the scene primitives which interact with that part of the screen) whose results can be used to accelerate the final per-pixel calculation.  In addition to yielding algorithmic benefits, this decomposition strategy maps well to parallel hardware.

@@

We'll start by analyzing a simple two-stage language called \ellStaged.  We say it is simple in that it supports boolean, integer, and product types, as well as let bindings, if expressions, and various primitive operations.  Notably, we leave out functions, fixed point operators, and full sum types, which will be added later.

The grammar for types, terms, values, and contexts in \ellStaged are given in Figure \ref{fig:coreSyntax}.  The first thing to notice is that stage $\bbone$ and stage $\bbtwo$ code are syntactically seperated at the type, term, and value levels, despite heavy overlap.  It would be tempting to merge the two stages into one syntactic class and maintain the stage invariants with the typing judgements.  Such a trick would reduce the total number of rules presented later, but at the cost of making those rules type derivation-directed, rather than syntax directed.  Alternatively, we chose to have one context merged accross both stages, rather than one context for each stage.  The intended effect of this decision is for bindings in one stage to shadow binding to the same variables in the other stage.  The reasons for this will become clear later.

@@

@@

The static semantics for \ellStaged~is given in Figure \ref{fig:coreStatics}.  Modulo the two-stage restriction and current absence of functions, it's based off of the system in \cite{davies96}, which was inspired by temporal logic (the source of our $\fut$ modality).  The static semantics comprises two judgements.  Those are:

@@

\begin{figure}
\caption{\ellStaged~Static Semantics}
\label{fig:coreStatics}
\begin{mathpar}
\infertypesone [\rmunit] 	{\ty {()}\rmunit}						{\cdot} \and
\infertypesone [int]			{\ty {i}{\rm int}}					{\cdot} \and
\infertypesone [bool]			{\ty {b} \rmbool}						{\cdot} \and
\infertypesone [hyp]			{\ty x A}								{x : A^\bbone \in \Gamma} \and
\infertypestwo [\rmunit]		{\ty {()}{\rmunit}}					{\cdot} \and
\infertypestwo [int]			{\ty {i}{\rm int}}					{\cdot} \and
\infertypestwo [bool]			{\ty {b} \rmbool}						{\cdot} \and
\infertypestwo [hyp]			{\ty x A}								{x : A^\bbtwo \in \Gamma} \and
\infertypesone [let]			{\ty {\letin{x}{e_1}{e_2}}{B}}		{\ty {e_1} A & \ty [\Gamma,x:A^\bbone] {e_2} B} \and
\infertypestwo [let]			{\ty {\letin{x}{e_1}{e_2}}{B}}		{\ty {e_1} A & \ty [\Gamma,x:A^\bbtwo] {e_2} B} \and
\infertypesone [\times I]	{\ty {(e_1,e_2)}{A\times B}}			{\ty {e_1} A & \ty {e_2} B} \and
\infertypesone [\times E_1]	{\ty {\pi_1~e} A}						{\ty e {A\times B}} \and
\infertypesone [\times E_2]	{\ty {\pi_2~e} B}						{\ty e {A\times B}} \and
\infertypestwo [\times I]	{\ty {(e_1,e_2)}{A\times B}}			{\ty {e_1} A & \typestwo {e_2} B} \and
\infertypestwo [\times E_1]	{\ty {\pi_1~e} A}						{\ty e {A\times B}} \and
\infertypestwo [\times E_2]	{\ty {\pi_2~e} B}						{\ty e {A\times B}} \and
\infertypesone [if] 			{\ty {\ifthen{e_1}{e_2}{e_3}} A}	{\ty {e_1} \rmbool & \ty {e_2} A & \ty {e_3} A} \and
\infertypestwo [if]			{\ty {\ifthen{e_1}{e_2}{e_3}} A}	{\ty {e_1} \rmbool & \ty {e_2} A & \ty {e_3} A} \and
\infertypesone [\fut I]		{\ty {\next~e}{\fut A}}				{\typestwo e A} \and
\infertypestwo [\fut E]		{\ty {\prev~e} A}						{\typesone e {\fut A}} \and
\end{mathpar}
\end{figure}

@@

The dynamics semantics for \ellStaged~is given in Figure \ref{fig:erasureSemantics} in big-step style.  Predictably, it comprises two judgements:

@@

\begin{figure}
\caption{\ellStaged~Erasure Dynamic Semantics}
\label{fig:erasureSemantics}
\begin{mathpar}
\inferreduceone [\rmunit]		{\red {()}{()}}							{\cdot} 											\and
\inferreduceone [int]				{\red {i}{i}}								{\cdot} 											\and
\inferreduceone [bool]			{\red {b}{b}}								{\cdot} 											\and
\inferreduceone [let]				{\red {\letin{x}{e_1}{e_2}}{v_2}}		{\sub [1] & \red{[v_1/x]e_2}{v_2}} 				\and
\inferreducetwo [\rmunit]		{\red {()}{()}}							{\cdot} 											\and
\inferreducetwo [int]				{\red {i}{i}}								{\cdot} 											\and
\inferreducetwo [bool]			{\red {b}{b}}								{\cdot} 											\and
\inferreducetwo [let]				{\red {\letin{x}{e_1}{e_2}}{v_2}}		{\sub [1] & \red{[v_1/x]e_2}{v_2}} 				\and
\inferreduceone [\times I]		{\red {(e_1,e_2)}{\valprod{v_1}{v_2}}}	{\sub [1] & \sub [2]} 							\and
\inferreduceone [\times E_1]		{\red {\pi_1~e}{v_1}}						{\red{e}{\valprod{v_1}{v_2}}} 					\and
\inferreduceone [\times E_2]		{\red {\pi_2~e}{v_2}}						{\red{e}{\valprod{v_1}{v_2}}} 					\and
\inferreducetwo [\times I]		{\red {(e_1,e_2)}{\valprod{v_1}{v_2}}}	{\sub [1] & \sub [2]}								\and
\inferreducetwo [\times E_1]		{\red {\pi_1~e}{v_1}}						{\red{e}{\valprod{v_1}{v_2}}} 					\and
\inferreducetwo [\times E_2]		{\red {\pi_2~e}{v_2}}						{\red{e}{\valprod{v_1}{v_2}}} 					\and
\inferreduceone [if_T] 			{\red {\ifthen{e_1}{e_2}{e_3}}{v}}		{\red{e_1}{\bf true}  & \red{e_2}{v}} 			\and
\inferreduceone [if_F] 			{\red {\ifthen{e_1}{e_2}{e_3}}{v}}		{\red{e_1}{\bf false} & \red{e_3}{v}}			\\
\inferreducetwo [if_T] 			{\red {\ifthen{e_1}{e_2}{e_3}}{v}}		{\red{e_1}{\bf true}  & \red{e_2}{v}} 			\and
\inferreducetwo [if_F] 			{\red {\ifthen{e_1}{e_2}{e_3}}{v}}		{\red{e_1}{\bf false} & \red{e_3}{v}}			\\
\inferreduceone [\fut I]			{\red {\next~e}{\valnext~v}}				{\reducetwosub}									\and
\inferreducetwo [\fut E]			{\red {\prev~e} v}						{\reduceone e {\valnext~v}} 
\end{mathpar}
\end{figure}

\begin{figure}
\caption{\ellStaged~Diagonal Semantics}
\label{fig:refSemantics}
\begin{mathpar}
\inferRedsplitone [\rmunit]		{\red {()}{()}{()}}												{\cdot} 												\and
\inferRedsplitone [int]			{\red {i}{i}{()}}													{\cdot} 												\and
\inferRedsplitone [bool]			{\red {b}{b}{()}}													{\cdot} 												\and
\inferRedsplitone [hyp]			{\red {[u,x]}{u}{x}}												{\cdot}													\and
\inferSpeculation [\rmunit]		{\red {()}{()}}													{\cdot} 												\and
\inferSpeculation [int]			{\red {i}{i}}														{\cdot} 												\and
\inferSpeculation [bool]			{\red {b}{b}}														{\cdot} 												\and
\inferSpeculation [hyp]			{\red {x}{x}}														{\cdot}													\and
\inferRedsplitone [let]			{\red {\letin{x}{e_1}{e_2}}{u_2}{\letin {x}{r_1}{r_2}}}		{\sub [1] & \red{[[u_1,x]/x]e_2}{u_2}{r_2}} 		\and
\inferSpeculation [let]			{\red {\letin{x}{e_1}{e_2}}{\letin{x}{r_1}{r_2}}}				{\sub [1] & \sub [2]}									\and
\inferRedsplitone [\times I]		{\red {(e_1,e_2)}{\valprod{u_1}{u_2}}{\valprod{r_1}{r_2}}}	{\sub [1] & \sub [2]}									\and
\inferRedsplitone [\times E_1]	{\red {\pi_1~e}{u_1}{\pi_1 r}}									{\red {e}{\valprod{u_1}{u_2}}{r}}					\and
\inferRedsplitone [\times E_2]	{\red {\pi_2~e}{u_2}{\pi_2 r}}									{\red {e}{\valprod{u_1}{u_2}}{r}} 					\and
\inferSpeculation [\times I]		{\red {(e_1,e_2)}{(r_1,r_2)}}									{\sub [1] & \sub [2]}									\and
\inferSpeculation [\times E_1]	{\red {\pi_1~e}{\pi_1 r}}										{\red {e}{r}}											\and
\inferSpeculation [\times E_2]	{\red {\pi_2~e}{\pi_2 r}}										{\red {e}{r}} 											\and
\inferRedsplitone [if_T] 		{\red {\ifthen{e_1}{e_2}{e_3}}{u}{(r_1;r_2)}}					{\red{e_1}{\bf true }{r_1} & \red{e_2}{u}{r_2}}	\and
\inferRedsplitone [if_F] 		{\red {\ifthen{e_1}{e_2}{e_3}}{u}{(r_1;r_3)}}					{\red{e_1}{\bf false}{r_1} & \red{e_3}{u}{r_3}}	\and
\inferSpeculation [if] 			{\red {\ifthen{e_1}{e_2}{e_3}}{\ifthen{r_1}{r_2}{r_3}}}		{\sub [1] & \sub [2] & \sub [3]}					\\
\inferRedsplitone [\fut I]		{\red {\next~e}{()}{r}}											{\rstwosub}											\and
\inferSpeculation [\fut E]		{\red {\prev~e}{r}}												{\rsone {e}{()}{r}}
\end{mathpar}
\end{figure}

@@

We now cover splitting, the process by which mixed code is teased apart into two individual chunks of code, one for each stage.  To begin, we first introduce the target lagnuage (for both stage), \ellTarget.   The grammar for \ellTarget~is given in figure \ref{fig:baseSyntax}.  We see that \ellTarget~is (as one would hope) a traditional single-stage language, with all the same basetypes as \ellStaged, in addition to products, if expressions, and full sums (injections and case expressions).  The presence of this last feature is a bit curious, since \ellStaged~did not support full sums.  The necessity for sums should become apparent in time.  The static and dynamic semantics of \ellTarget~are not given, but should be obvious.  They are given by the judgements:

@@

@@

\begin{figure*}
\caption{Term Splitting}
\label{fig:termSplit}
\begin{mathpar}
\infersplitone [unit]			{\spl {()}{\rmunit}{((),()),\_.()}}															{\cdot} \and
\infersplittwo [unit]			{\spl {()}{\rmunit}{(),\_.()}}																{\cdot} \and
\infersplitone [int]			{\spl {i}{\rm int}{(i,()),\_.()}}															{\cdot} \and
\infersplittwo [int]			{\spl {i}{\rm int}{(),\_.i}}																	{\cdot} \and
\infersplitone [\times I] 	{\spl {(e_1,e_2)}{A\times B}
									{ \begin{array}{l}
									\left(
										\letin{y_1}{c_1}{\letin{y_2}{c_2}{
										((\pi_1 y_1, \pi_1 y_2),(\pi_2 y_1, \pi_2 y_2))}}
									\right), \\
									l.(\projbind 1,\projbind 2)
									\end{array}}}																				{\sub 1 A & \sub 2 B} \and
\infersplittwo [\times I]	{\spl {(e_1,e_2)}{A\times B}{(p_1,p_2),l.(\projbind 1,\projbind 2)}}					{\sub 1 A & \sub 2 B} \and
\infersplitone [\times E_1]	{\spl {\pi_1~e}{A}{\letin{y}{c}{(\pi_1(\pi_1 y),\pi_2 y)},l.\pi_1~r}}					{\sub {} {A\times B}} \and
\infersplitone [\times E_2]	{\spl {\pi_2~e}{B}{\letin{y}{c}{(\pi_2(\pi_1 y),\pi_2 y)},l.\pi_2~r}}					{\sub {} {A\times B}} \and
\infersplittwo [\times E_1]	{\spl {\pi_1~e}{A}{p,l.\pi_1~r}}															{\sub {} {A\times B}} \and
\infersplittwo [\times E_2]	{\spl {\pi_2~e}{B}{p,l.\pi_2~r}}															{\sub {} {A\times B}} \and
\infersplitone [hyp]			{\spl {v}{A}{(x,()),\_.()}}																	{x:A^\bbone \in \Gamma} \and
\infersplittwo [hyp]			{\spl {v}{A}{(),\_.x}}																		{x:A^\bbtwo \in \Gamma} \\
\infersplitone [let]			{\spl {\letin{x}{e_1}{e_2}}{B}{
								\begin{array}{l}
									\left(\talllet{y_1}{c_1}{
									\talllet{y_2}{(\letin{x}{\pi_1 y_1}{c_2})}{
									(\pi_1 y_2, (\pi_2 y_1, \pi_2 y_2))}}\right), \\
									l.\letin{x}{(\projbind 1)}{\projbind 2}
								\end{array}}}																					{\sub 1 A & \sub [\Gamma,x:A^\bbone] 2 B} \and
\infersplittwo [let]			{\spl {\letin{x}{e_1}{e_2}}{B}{(p_1,p_2), l.\letin{x}{(\projbind 1)}{\projbind 2}}}	{\sub 1 A & \sub [\Gamma,x:A^\bbtwo] 2 B} \and
\infersplitone [\fut I]		{\spl {\next~e}{\fut A}{((),p),l.r}}														{\splittwosub {} A} \and
\infersplittwo [\fut E]		{\spl {\prev~e}{A} {\pi_2~c,l.r}}															{\splitonesub {} {\fut A}} \and
\infersplitone [if]			{\spl {\left( \tallif {e_1}{e_2}{e_3} \right)}{A} 
								{\begin{array}{l}
									\left(\talllet{z}{c_1}{\tallif {\pi_1 z}
										{\letin{y}{c_2}{(\pi_1 y, (\pi_2 z, \inl(\pi_2 y)))}}
										{\letin{y}{c_3}{(\pi_1 y, (\pi_2 z, \inr(\pi_2 y)))}}}
									\right), \\
									l.(\projbind 1;\caseof{\pi_2 l}{l_2.r_2}{l_3.r_3})
								\end{array}}}																					{\sub 1 \rmbool & \sub 2 A & \sub 3 A } \and
\infersplittwo [if]			{\spl {\left(\tallif {e_1}{e_2}{e_3}\right)}{A} 
								{(p_1,p_2,p_3),l.\left(\tallif{(\projbind 1)}{(\projbind 2)}{(\projbind 3)}\right)}}	{\sub 1 \rmbool & \sub 2 A & \sub 3 A } \and
\end{mathpar}
\end{figure*}

@@

\section {Adding Full Sums and Functions}

In this section, we consider adding two more features to our source language: sums and functions.  We will call this augmented language \lamStaged.  The grammar for \lamStaged is given in \ref{fig:lamStagedSyntax}.
